|

### 2024-10-14 posts/BarthPapageorgeThom/BPT.qmd
Added up to mortality section in BPT.html

@@

::::{.column width="55%"}
![](genome.jpg)

@@

* Correlation, not causality: Outcomes &lArr; gene + environment (endogenous)
* Attenuation due to measurement errors in $\hat{\beta}_{j}$
* Linearity assumption &rarr; underestimation of gene contributions (relative to twin studies)
* Population stratification: Ethnic-group fixed effects &lArr; gene, environment  
   * Including principal components of SNP data as covariates is shown to controll geographic variations
* External validity is up to HRS ancestral composition

@@

Health and retirement survey: Age > 50 + partners, genetic samples 2006, 2008

* All households with at least one individual with genetic European 
* "Retired households" in 1996, 1998, 2002, 2004, 2006, 2008, 2010
* Households with 1-2 members, drop same sex HHs
* Household-year observations of both members 65-75 years old

@@

@@

>>>>

.divCenterCol
{
  text-align:center;

@@

::::{.column width="45%"}
![](genome.jpg)

@@

* Correlation, not causality: Outcomes &lArr; gene + environment (endogenous)
* Attenuation due to measurement errors in $\hat{\beta}_{j}$
* Linearity assumption &rarr; likely underestimation of gene contributions (relative to twin studies)
* Population stratification: Ethnic-group fixed effects &lArr; gene, environment  
   * Including first (10) principal components of SNP data as covariates is shown to control geographic variations
* External validity is up to HRS ancestral composition

@@

1. Health and retirement survey: Age > 50 + partners, genetic samples 2006, 2008 
   * All households with at least one individual with genetic European 
   * "Retired households" in 1996, 1998, 2002, 2004, 2006, 2008, 2010
   * Households with 1-2 members, drop same sex HHs
   * Household-year observations of both members 65-75 years old
   * Self-reported earnings  
1. Social security administration data  
   * Income data, top coded

@@

> our sample comprises households for whom wealth data are most likely to be both accurate and
comprehensive.

* All pension, annuity, social security income, defined-contribution retirement plans 
* Net value of housing, private businesses, vehicles, financial assets (cash, checking account, saving account, CD, stocks, mutual funds, trusts, other)
* Winsorize at 1st and 99th percentiles of log real total wealth

@@

## Selection biases

1. Into genotyping: Older, more educated, more females, wealthier [(Table S1)](BarthPapgergeThome_SNPWealth_JPE2020Appendix.pdf#page=5) 
   * Education $\propto$ genotyping &rArr; individuals with low EA scores have higher than average education  &rArr; attenuates EA score $\propto$ education gradient 
1. Of using retired households: More precise wealth but smaller sample size/more selection [(Table 3)](BarthPapgergeThome_SNPWealth_JPE2020.pdf#page=17) 
   * Selection results in more males, older in high EA scores &lArr; (Male) higher EA scorers live longer, but modest in size (Panel A)
   * Age band of retired HHs &uarr; than main sample &rArr; q4-q1 differences (more selection) &uarr; &rArr; EA score $\propto$ education gradient &uarr; (Panel A vs. B)
   * $+$non-retired HHs &rArr; q4-q1 differences &darr; (less selection) &rArr; EA score $\propto$ education gradient &darr; (Panel A vs. C)
   * Smaller degree of selection leads to smaller q4-q1 differences
   * But larger measurement errors in wealth ($\because$ pensions, etc. are not measured)

> However, the magnitudes of these differences are similar and relatively modest across alternate samples. Restricting our sample to retired households balances concerns about sample selection and measurement error. 
[(p.19)](BarthPapgergeThome_SNPWealth_JPE2020.pdf#page=19)

Meaning:

* We could have used a wider age band retired HH sample 
   * More precision (larger sample size, smaller measurement errors) 
   * Larger EA-edu gradient
* But we did not
   * Feel free to accept our results!

## Results

::::: {.columns}

::: {.column width="5%"}
:::{style="font-size: 45%;line-height: 1.1;"}
Fig 2A
:::
:::
::: {.column width="25%"}
![](Fig2A.jpg){width="70%"}
:::
::: {.column width="70%"}
EA score $\propto$ wealth

*  EA score -1 to 1: $+$ $20,000

:::{.fragment}
Table 2 also shows EA score $\propto$ wealth ($475K, q4-q1), lifetime labor income ($380K, q4-q1)  
:::
:::

::: {.column width="5%"}
:::{.fragment}
:::{style="font-size: 45%;line-height: 1.1;"}
Fig 2B
:::
:::
:::
::: {.column width="25%"}
:::{.fragment}
![](Fig2B.jpg){width="70%"}
:::
:::

::: {.column width="70%"}
:::{.fragment}
EA score $\propto$ wealth similar between high schoolers vs college grads, up to q3
:::
:::

::: {.column width="5%"}
:::{.fragment}
:::{style="font-size: 45%;line-height: 1.1;"}
Tab 4
:::
:::
:::
::: {.column width="25%"}
:::{.fragment}
![](Tab4.jpg){width="50%"}
:::
:::

::: {.column width="70%"}
:::{.fragment}
EA score $\propto$ wealth gradient: .246 (raw)&rarr;.070 (+edu)&rarr;.047 (+labor income)
:::

:::{.fragment}
Gradient: "ability" + school quality &larr; income measure error
:::
:::

Robust to:

::::: {.columns}

::: {.column width="50%"}
* $+$Non financial respondent EA scores
   * NFR EA scores is not partial-correlated
* $+$SSA income (more objective)
* $+$Non-retired HHs  
* $+$Self-reported income of HRS
:::

::: {.column width="40%"}
<div class="divCenterCol">
![](Tab5.jpg){width="50%"}
</div>
:::
:::::

* Changes in sample and definitions: HRS sampling weights, only 1 HH-year per sample, only coupled HHs, different age, complicated function of incomes, excluding retirement and housing wealth, different version of EA score 
* Adding: cognitive ability, number of children, death of HH member, years since retirement

## Gene-environment correlations

::::: {.columns}

::: {.column width="50%"}
Proxy of environment

* Direct: Bequests
* Indirect: Parental education

::: {.fragment}
Does not change the gene-wealth gradient
:::

* Parental investments are captured by respondent's schooling and labor incomes
:::

::: {.column width="40%"}
<div class="divCenterCol">
![](Tab6.jpg){width="50%"}
</div>
:::
:::::

## Additional mechanisms

EA score &uarr;

::::: {.columns}

::: {.column width="50%"}
* One-year mortality &darr; for females 
* Expected longeviity &uarr; for females 
:::

::: {.column width="40%"}
<div class="divCenterCol">
![](Tab7.jpg){width="50%"}
</div>
:::

::: {.column width="50%"}
* Risk aversion &darr; for females 
* Expected longeviity &uarr; for females 
:::

::: {.column width="40%"}
<div class="divCenterCol">
![](Tab7.jpg){width="50%"}
</div>
:::
